Statistically-estimated tree composition for the northeastern United States \\ at the time of European settlement

### 2015-09-05 paperPlos.tex
added arxiv and plos versions of tex

\documentclass[12pt]{article}
\usepackage[]{graphicx}
\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
%\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
%\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
%\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
%\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
%\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
%\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
%\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
%\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
%\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

%\usepackage{framed}
%\makeatletter
%\newenvironment{kframe}{%
% \def\at@end@of@kframe{}%
% \ifinner\ifhmode%
%  \def\at@end@of@kframe{\end{minipage}}%
%  \begin{minipage}{\columnwidth}%
% \fi\fi%
% \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
% \colorbox{shadecolor}{##1}\hskip-\fboxsep
%     % There is no \\@totalrightmargin, so:
%     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
% \MakeFramed {\advance\hsize-\width
%   \@totalleftmargin\z@ \linewidth\hsize
%   \@setminipage}}%
% {\par\unskip\endMakeFramed%
% \at@end@of@kframe}
%\makeatother

%\definecolor{shadecolor}{rgb}{.97, .97, .97}
%\definecolor{messagecolor}{rgb}{0, 0, 0}
%\definecolor{warningcolor}{rgb}{1, 0, 1}
%\definecolor{errorcolor}{rgb}{1, 0, 0}
%\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\usepackage{array}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[authoryear]{natbib}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newcommand{\lyxaddress}[1]{
\par {\raggedright #1
\vspace{1.4em}
\noindent\par}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{times}
\usepackage{graphics,natbib}

@@

\makeatother

\begin{document}

\title{Statistically-estimated tree composition for the northeastern United
States \\
at the time of Euro-American settlement}

\author{Christopher J. Paciorek\textsuperscript{1,{*}}, Simon J. Goring\textsuperscript{2,\&},
Andrew L. Thurman\textsuperscript{3,\&}, \\
Charles V. Cogbill\textsuperscript{4}, John W. Williams\textsuperscript{2,3},
David J. Mladenoff\textsuperscript{6}, \\
Jody A. Peters\textsuperscript{7}, Jun Zhu\textsuperscript{8}, and
Jason S. McLachlan\textsuperscript{7}}

\maketitle

\lyxaddress{\textsuperscript{1}Department of Statistics, University of California, Berkeley, California, USA}

\lyxaddress{\textsuperscript{2}Department of Geography, University of Wisconsin, Madison, Wisconsin, USA}

\lyxaddress{\textsuperscript{3}Department of Statistics, University of Iowa, Iowa City, Iowa, USA}

\lyxaddress{\textsuperscript{4}Harvard Forest, Harvard University, Petersham, Massachusetts, USA}

\lyxaddress{\textsuperscript{5}Center for Climatic Research, University of Wisconsin,
Madison, Wisconsin, USA}

\lyxaddress{\textsuperscript{6}Department of Forestry and Wildlife Ecology,
University of Wisconsin, Madison, Wisconsin, USA}

\lyxaddress{\textsuperscript{7}Department of Biological Sciences, University
of Notre Dame, Notre Dame, Indiana, USA}

\lyxaddress{\textsuperscript{8}Department of Statistics, University of Wisconsin,
Madison, Wisconsin, USA}

\lyxaddress{\textsuperscript{{*}}Corresponding author; E-mail: paciorek@stat.berkeley.edu}

\lyxaddress{\textsuperscript{\&}These authors contributed equally to this work.
\newpage}

\begin{abstract}
We present a gridded 8 km-resolution data product of the estimated
composition of tree taxa at the time of Euro-American settlement of
the northeastern United States and the statistical methodology used
to produce the product from trees recorded by land surveyors. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height for 22 tree taxa, generally at the genus
level. The data come from settlement-era public survey records that
are transcribed and then aggregated spatially, giving count data.
The domain is divided into two regions, eastern (Maine to Ohio) and
midwestern (Indiana to Minnesota). Public Land Survey point data in
the midwestern region (ca. 0.8-km resolution) are aggregated to a
regular 8 km grid, while data in the eastern region, from Town Proprietor
Surveys, are aggregated at the township level in irregularly-shaped
local administrative units. The product is based on a Bayesian statistical
model fit to the count data that estimates composition on a regular
8 km grid across the entire domain. The statistical model is designed
to handle data from both the regular grid and the irregularly-sized
townships and allows us to estimate composition at locations with
no data and to smooth over noise caused by limited counts in locations
with data. Critically, the model also allows us to quantify uncertainty
in our composition estimates, making the product suitable for applications
employing data assimilation. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale Euro-American settlement. In addition
to specific regional questions, the data product can also serve as
a baseline against which to investigate how forests and ecosystems
change after intensive settlement. The data product is being made
available at the NIS data portal as version 0.4 under the CC BY license.

Keywords: biogeography, species composition, old-growth forests, spatial
modeling, Bayesian statistical model, vegetation mapping
\end{abstract}

\section{Introduction}

@@

Euro-American settlement and subsequent land use change occurred in
a time-transient fashion across North America and were accompanied
by land surveys needed to demarcate land for land tenure and use.
Various systems were used by surveyors to locate legal boundary markers,
usually by recording and marking trees adjacent to survey markers.
These data provide vegetation information that can be mapped and used
quantitatively to represent the period of settlement. Early surveys
(from 1620 until 1825) in the northeastern United States provide spatially-aggregated
data at the township level \citep{Cogb:etal:2002,thompson2013four},
with typical township size on the order of 200 $\mbox{km}^{2}$ and
no information about the locations of individual trees; we refer to
these as the Town Proprietor Survey (TPS). Later surveys after the
establishment of the U.S. Public Land Survey System (PLS) by the General
Land Office (GLO) provide point-level data along a regular grid, with
one-half mile (800 m) spacing, for Ohio and westward during the period
1785 to 1907 \citep{bourdo1956review,pattison1957beginnings,schulte2001original,goring2015composition}.
At each point 2-4 trees were identified, and the common name, diameter
at breast height, and distance and bearing from the point were recorded.
Survey instructions during the PLS varied through time and by point
type. Accounting for this variation requires data screening to maximize
consistency among points and the application of spatially-varying
correction factors \citep{cogbill2015corrections,goring2015composition}
to accurately assess tree stem density, basal area and biomass from
the early settlement records, but the impact on composition estimates
is limited \citep{liu2011broadscale}. Surveyors sometimes used ambiguous
common names, which requires matching to scientific names and standardization
\citep{mladenoff2002narrowing,goring2015composition}.

Logging, agriculture, and land abandonment have left an indelible
mark on forests in the northeastern United States \citep{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
However most studies have assessed these effects in individual states
or smaller domains \citep{friedman2005regional,rhemtulla2009historical}
and with various spatial resolutions, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \citet{goring2015composition}
provide a new dataset of forest composition, biomass, and stem density
based on PLS data for the upper Midwest that is resolved to an 8 km
by 8 km grid cell scale, providing broad spatial coverage at a spatial
scale that can be compared to modern forests using Forest Inventory
and Analysis products \citep{gray2012forest}. Combined with additional,
coarsely-sampled PLS data from Illinois and Indiana, newly-digitized
data from southern Michigan, and with the TPS data, this gives us
raw data for much of the northeastern United States. However, there
are several limitations of using the raw data that can be alleviated
by the use of a statistical model to develop a statistically-estimated
data product. First, the PLS and TPS data only provide estimates of
within-cell variance that do not account for information from nearby
locations. Second, there are data gaps: the available digitized data
from Illinois and Indiana represent a small fraction of those states
and missing townships are common in the TPS data. Third, the TPS and
PLS data have fundamentally different sampling design and spatial
resolution. Our statistical model allows us to provide a spatially-complete
data product of settlement-era tree composition for a common 8 km
grid with uncertainty across the northeastern U.S.

In Section \ref{sec:Data} we describe the data sources, while Section
\ref{sec:Statistical-model} describes our statistical models. In
Section \ref{sec:Model-comparison} we quantitatively compare competing
statistical specifications, and in Section \ref{sec:Data-product}
we describe the final data product. In Section \ref{sec:Discussion}
we discuss limitations of the statistical model, and we list related
data products under development.

\section{Data\label{sec:Data}}

The raw data were obtained from land division survey records collated
and digitized from across the northeastern U.S. by a number of researchers
(Fig. \ref{fig:Spatial-domain}). For the states of Minnesota, Wisconsin,
Illinois, Indiana, and Michigan (the midwestern subdomain), digitized
data are available at PLS survey point locations and have been aggregated
to a regular 8 km grid in the Albers projection. (Note that for Indiana
and Illinois, at the moment trees are associated with township centroids
and then assigned to 8 km grid cells based on the centroid but in
the near future we will have point locations available for each tree.)
For the states of Ohio, Pennsylvania, New Jersey, New York and the
six New England states (the eastern subdomain), data are aggregated
at the township level. There are also data from a single township
in Quebec and a single township in northern Delaware. Digitization
of PLS data in Minnesota, Wisconsin and Michigan is essentially complete,
with PLS data for nearly all 8 km grid cells, but data in Illinois
and Indiana represent a sample of the full set of grid cells, with
survey record transcription ongoing. Data for the eastern states are
available for a subset of the full set of townships covering the domain;
the TPS data for some townships were lost, incomplete, or have not
been located \citep{Cogb:etal:2002}.

\begin{figure}
\label{fig:domain}\includegraphics[scale=1.3]{fig1}

\caption{Spatial domain of the northeastern United States, with locations with
data shown in gray. Locations are grid cells in midwestern portion
and townships in eastern portion. In addition to locations without
data being indicated in white, grid cells completely covered in water
are white (e.g., a few locations in the northwestern portion of the
domain in the states of Minnesota and Wisconsin).\label{fig:Spatial-domain}}
\end{figure}

Note that surveys occurred over a period of more than 200 years as
European colonists (before U.S. independence) and the United States
settled what is now the northeastern and midwestern United States.
Our estimates are for the period of settlement represented by the
survey data and therefore are time-transgressive; they do not represent
any single point in time across the domain, but rather the state of
the landscape at the time just prior to widespread Euro-American settlement
and land use \citep{Whit:1996,Cogb:etal:2002}. These forest composition
datasets do include the effects of Native American land use and early
Euro-American settlement activities \citep[e.g.,][]{Blac:etal:2006},
but it is likely that the imprint of this earlier land use is highly
concentrated rather than spatially extensive \citep{munoz2014defining}.

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available in \cite{goring2015composition};
key elements include the use of only corner points, the use of only
the two closest trees at each corner point, spatially-varying correction
factors for sampling effort, and a standardized taxonomy table. The
lower Midwest (Illinois, Indiana) data were purchased from the Indiana
State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. Digitization of the Illinois and Indiana data is still underway,
so many grid cells contained no data at the time the statistical model
was fit. Note that the number of trees per grid cell varies depending
on the number of survey points in a cell, with an average of 124 trees
per cell. The gridded data at the 8 km resolution for the upper Midwest
are available through the NIS Data portal (accession number and version
to be inserted when available). The TPS data were compiled by C.V.
Cogbill from a myriad of archival sources representing land division
surveys conducted in connection with local settlement.

@@

Diameters are only recorded in the PLS data. Although surveyors avoided
using small trees, there was no consistent lower diameter limit. The
PLS data generally represent trees greater than 8 inches (\textasciitilde{}20
cm) diameter at breast height (dbh), but with some trees as small
as 1 inch dbh (smaller trees were much more common in far northern
Minnesota). TPS data have no information about dbh, but the trees
were large enough to blaze and are presumed to be relatively large
trees useful for marking property boundaries.

@@

\section{Statistical model\label{sec:Statistical-model}}

@@

\subsection{Data model}

We start by describing the basic model for those states for which
we have raw data on the 8 km grid, and in Section \ref{sub:Model-for-township}
we describe the extension of the model to accommodate data aggregated
at the township level.

The statistical model treats the observations as coming from a multinomial
distribution with a (latent) vector of proportions for each grid cell,
\[
y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i})),
\]
where $y_{i}$ is the vector of counts for the $P$ taxa at the $i$th
cell, $n_{i}$ is the number of trees counted in the cell, and $\theta(s_{i})$
is the vector of unknown proportions for those taxa at that cell.
Note that we use a standard multinomial distribution without overdispersion,
because the set of trees in the dataset is roughly uniformly sampled
across the cells or townships \citep{goring2015composition}.

The proportions, $\theta_{p}(s_{i}),\, p=1,\ldots,P$, are modeled
spatially by a set of $P$ Gaussian spatial processes, one per taxon,
$\alpha_{p}(s_{i}),\, p=1,\ldots,P$. This collection of processes
defines a multivariate spatial process for composition. The $\alpha_{p}(s)$
processes are defined on the 8 km grid, $\alpha_{p}=\{\alpha_{p}(s_{1}),\ldots,\alpha_{p}(s_{m})\}$
for the $m$ grid cells. In Section \ref{sub:Latent-Variable-Model}
we introduce a multinomial probit model that relates the $\alpha_{p}(s)$
processes to the proportion processes, $\theta_{p}(s)$, via the introduction
of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.

@@

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \citep{Bane:etal:2004} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with Matern
covariance \citep{Lind:etal:2011}.

\subsection{Spatial process models}

MRF models work directly with the precision matrix of the values of
the spatial process, so calculation of the prior density of $\alpha_{p}$
is computationally simple \citep{Rue:Held:2005}, but in situations
where the likelihood is not normal, it can be difficult to set up
effective MCMC algorithms that are able to move in the high-dimensional
space of $\alpha_{p}$. The latent variable representation of Section
\ref{sub:Latent-Variable-Model} helps to alleviate this problem.
Next we describe two alternative spatial models that we considered;
in Section \ref{sec:Model-comparison}, we use cross-validation to
choose between these two representations.

@@

Gaussian processes (GP) are also standard models for spatial processes \citep{Bane:etal:2004}.
GP models are computationally challenging for large datasets because
of manipulations involving large covariance matrices. Given this,
\cite{Lind:etal:2011} proposed a new framework for using Gaussian
MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
partial differential equations (SPDEs).

>>>>

% Template for PLoS
% Version 3.1 February 2015
%
% To compile to pdf, run:
% latex plos_template
% bibtex plos_template
% latex plos_template
% latex plos_template
% dvipdf plos_template
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig." instead of "Figure".
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment. For example, CO2 will be CO\textsubscript{2}.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% rotating package for sideways tables
\usepackage{rotating}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

% Use the PLoS provided BiBTeX style
\bibliographystyle{latex/plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{\includegraphics[width=2.0in]{latex/PLOS-Submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% user-provided packages
\usepackage{array}

%% Include all macros below

@@

%% END MACROS SECTION

\begin{document}
\vspace*{0.35in}

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Statistically-estimated Tree Composition for the\\
Northeastern United States at Euro-American Settlement}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Christopher J. Paciorek\textsuperscript{1,*},
Simon J. Goring\textsuperscript{2,\Yinyang},
Andrew L. Thurman\textsuperscript{3,\Yinyang},
Charles V. Cogbill\textsuperscript{4},
John W. Williams\textsuperscript{2,5},
David J. Mladenoff\textsuperscript{6},
Jody A. Peters\textsuperscript{7},
Jun Zhu\textsuperscript{8},
Jason S. McLachlan\textsuperscript{7,\ddag}
\\
\bigskip
\bf{1} Department of Statistics, University of California, Berkeley, California,
USA
\\
\bf{2} Department of Geography, University of Wisconsin, Madison, Wisconsin,
USA
\\
\bf{3} Department of Statistics, University of Iowa, Iowa City, Iowa, USA
\\
\bf{4} Harvard Forest, Harvard University, Petersham, Massachusetts, USA
\\
\bf{5} Center for Climatic Research, University of Wisconsin, Madison,
Wisconsin, USA
\\
\bf{6} Department of Forestry and Wildlife Ecology, University of Wisconsin,
Madison, Wisconsin, USA
\\
\bf{7} Department of Biological Sciences, University of Notre Dame, Notre
Dame, Indiana, USA
\\
\bf{8} Department of Statistics, University of Wisconsin, Madison, Wisconsin,
USA
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
 \ddag PI of the PalEON Project

% Current address notes
% \textcurrency a Insert current address of first author with an address update
% \textcurrency b Insert current address of second author with an address update
% \textcurrency c Insert current address of third author with an address update

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* paciorek@stat.berkeley.edu (CJP)

%TODO for PLOS\_ONE

%* coverletter
%* general guideline

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}

We present a gridded 8 km-resolution data product of the estimated
composition of tree taxa at the time of Euro-American settlement of
the northeastern United States and the statistical methodology used
to produce the product from trees recorded by land surveyors. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height for 22 tree taxa, generally at the genus
level. The data come from settlement-era public survey records that
are transcribed and then aggregated spatially, giving count data.
The domain is divided into two regions, eastern (Maine to Ohio) and
midwestern (Indiana to Minnesota). Public Land Survey point data in
the midwestern region (ca. 0.8-km resolution) are aggregated to a
regular 8 km grid, while data in the eastern region, from Town Proprietor
Surveys, are aggregated at the township level in irregularly-shaped
local administrative units. The product is based on a Bayesian statistical
model fit to the count data that estimates composition on the
8 km grid across the entire domain. The statistical model is designed
to handle data from both the regular grid and the irregularly-sized
townships and allows us to estimate composition at locations with
no data and to smooth over noise caused by limited counts in locations
with data. Critically, the model also allows us to quantify uncertainty
in our composition estimates, making the product suitable for applications
employing data assimilation. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale Euro-American settlement. In addition
to specific regional questions, the data product can also serve as
a baseline against which to investigate how forests and ecosystems
change after intensive settlement. The data product is being made
available at the NIS data portal as version 0.4.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\linenumbers

\section*{Introduction}

@@

Euro-American settlement and subsequent land use change occurred in
a time-transient fashion across North America and were accompanied
by land surveys needed to demarcate land for land tenure and use.
Various systems were used by surveyors to locate legal boundary markers,
usually by recording and marking trees adjacent to survey markers.
These data provide vegetation information that can be mapped and used
quantitatively to represent the period of settlement. Early surveys
(from 1620 until 1825) in the northeastern United States provide spatially-aggregated
data at the township level \cite{Cogb:etal:2002,thompson2013four},
with typical township size on the order of 200 km\textsuperscript{2} and
no information about the locations of individual trees; we refer to
these as the Town Proprietor Survey (TPS). Later surveys after the
establishment of the U.S. Public Land Survey System (PLS) by the General
Land Office (GLO) provide point-level data along a regular grid, with
one-half mile (800 m) spacing, for Ohio and westward during the period
1785 to 1907 \cite{bourdo1956review,pattison1957beginnings,schulte2001original,goring2015composition}.
At each point 2-4 trees were identified, and the common name, diameter
at breast height, and distance and bearing from the point were recorded.
Survey instructions during the PLS varied through time and by point
type. Accounting for this variation requires data screening to maximize
consistency among points and the application of spatially-varying
correction factors \cite{goring2015composition}
to accurately assess tree stem density, basal area and biomass from
the early settlement records, but the impact on composition estimates
is limited \cite{liu2011broadscale}. Surveyors sometimes used ambiguous
common names, which requires matching to scientific names and standardization
\cite{mladenoff2002narrowing,goring2015composition}.

Logging, agriculture, and land abandonment have left an indelible
mark on forests in the northeastern United States \cite{foster1998land,rhemtulla2009legacies,thompson2013four,goring2015composition}.
However most studies have assessed these effects in individual states
or smaller domains \cite{friedman2005regional,rhemtulla2009historical}
and with various spatial resolutions, from townships (36 square miles)
to forest zones of hundreds or thousands of square miles. \cite{goring2015composition}
provide a new dataset of forest composition, biomass, and stem density
based on PLS data for the upper Midwest that is resolved to an 8 km
by 8 km grid cell scale, providing broad spatial coverage at a spatial
scale that can be compared to modern forests using Forest Inventory
and Analysis products \cite{gray2012forest}. Combined with additional,
coarsely-sampled PLS data from Illinois and Indiana, newly-digitized
data from southern Michigan, and with the TPS data, this gives us
raw data for much of the northeastern United States. However, there
are several limitations of using the raw data that can be alleviated
by the use of a statistical model to develop a statistically-estimated
data product. First, the PLS and TPS data only provide estimates of
within-cell variance that do not account for information from nearby
locations. Second, there are data gaps: the available digitized data
from Illinois and Indiana represent a small fraction of those states
and missing townships are common in the TPS data. Third, the TPS and
PLS data have fundamentally different sampling design and spatial
resolution. Our statistical model allows us to provide a spatially-complete
data product of settlement-era tree composition for a common 8 km
grid with uncertainty across the northeastern U.S.

In \emph{Data} we describe the data sources, while in \emph{Statistical model}
we describe the statistical model used to create the data product.
In \emph{Model comparison} we quantitatively compare competing
statistical specifications, and \emph{Data Product}
we describe the final data product. In \emph{Discussion}
we discuss limitations of the statistical model, and we list related
data products under development.

\section*{Methods}

\subsection*{Data\label{sec:Data}}

The raw data were obtained from land division survey records collated
and digitized from across the northeastern U.S. by a number of researchers
(Fig. \ref{fig:Spatial-domain}). For the states of Minnesota, Wisconsin,
Illinois, Indiana, and Michigan (the midwestern subdomain), digitized
data are available at PLS survey point locations and have been aggregated
to a regular 8 km grid in the Albers projection. (Note that for Indiana
and Illinois, at the moment trees are associated with township centroids
and then assigned to 8 km grid cells based on the centroid but in
the near future we will have point locations available for each tree.)
For the states of Ohio, Pennsylvania, New Jersey, New York and the
six New England states (the eastern subdomain), data are aggregated
at the township level. There are also data from a single township
in Quebec and a single township in northern Delaware. Digitization
of PLS data in Minnesota, Wisconsin and Michigan is essentially complete,
with PLS data for nearly all 8 km grid cells, but data in Illinois
and Indiana represent a sample of the full set of grid cells, with
survey record transcription ongoing. Data for the eastern states are
available for a subset of the full set of townships covering the domain;
the TPS data for some townships were lost, incomplete, or have not
been located \cite{Cogb:etal:2002}.

\begin{figure}
% \label{fig:domain}\includegraphics[scale=1.3]{fig1.pdf}

\caption{{\bf Spatial domain of the northeastern United States, with locations with
data shown in gray.} \\ Locations are grid cells in midwestern portion
and townships in eastern portion. In addition to locations without
data being indicated in white, grid cells completely covered in water
are white (e.g., a few locations in the northwestern portion of the
domain in the states of Minnesota and Wisconsin).}}
\label{fig:Spatial-domain}
\end{figure}

Note that surveys occurred over a period of more than 200 years as
European colonists (before U.S. independence) and the United States
settled what is now the northeastern and midwestern United States.
Our estimates are for the period of settlement represented by the
survey data and therefore are time-transgressive; they do not represent
any single point in time across the domain, but rather the state of
the landscape at the time just prior to widespread Euro-American settlement
and land use \cite{Whit:1996,Cogb:etal:2002}. These forest composition
datasets do include the effects of Native American land use and early
Euro-American settlement activities, e.g. \cite{Blac:etal:2006},
but it is likely that the imprint of this earlier land use is highly
concentrated rather than spatially extensive \cite{munoz2014defining}.

Extensive details on the upper Midwest (Minnesota, Wisconsin, Michigan)
data and processing steps are available \cite{goring2015composition};
key elements include the use of only corner points, the use of only
the two closest trees at each corner point, spatially-varying correction
factors for sampling effort, and a standardized taxonomy table. The
lower Midwest (Illinois, Indiana) data were purchased from the Indiana
State Archives (Indiana) and Hubtack Document Resources (hubtack.com;
Illinois) and processed using similar steps as for the upper Midwest
data. Digitization of the Illinois and Indiana data is still underway,
so many grid cells contained no data at the time the statistical model
was fit. Note that the number of trees per grid cell varies depending
on the number of survey points in a cell, with an average of 124 trees
per cell. The gridded data at the 8 km resolution for the upper Midwest
are available through the NIS Data portal (accession number and version
to be inserted when available). The TPS data were compiled by C.V.
Cogbill from a myriad of archival sources representing land division
surveys conducted in connection with local settlement.

@@

Diameters are only recorded in the PLS data. Although surveyors avoided
using small trees, there was no consistent lower diameter limit. The
PLS data generally represent trees greater than 8 inches (ca. 20
cm) diameter at breast height (dbh), but with some trees as small
as 1 inch dbh (smaller trees were much more common in far northern
Minnesota). TPS data have no information about dbh, but the trees
were large enough to blaze and are presumed to be relatively large
trees useful for marking property boundaries.

@@

\subsection*{Statistical model\label{sec:Statistical-model}}

@@

\paragraph*{Data model}

We start by describing the basic model for those states for which
we have raw data on the 8 km grid, and in \emph{Model for township data}
we describe the extension of the model to accommodate data aggregated
at the township level.

The statistical model treats the observations as coming from a multinomial
distribution with a (latent) vector of proportions for each grid cell,
\[
y_{i}\sim\mbox{Multi}(n_{i},\theta(s_{i})),
\]
where $y_{i}$ is the vector of counts for the \emph{P} taxa at the \emph{i}th
cell, $n_{i}$ is the number of trees counted in the cell, and $\theta(s_{i})$
is the vector of unknown proportions for those taxa at that cell.
Note that we use a standard multinomial distribution without overdispersion,
because the set of trees in the dataset is roughly uniformly sampled
across the cells or townships \cite{goring2015composition}.

The proportions, $\theta_{p}(s_{i}),\, p=1,\ldots,P$, are modeled
spatially by a set of $P$ Gaussian spatial processes, one per taxon,
$\alpha_{p}(s_{i}),\, p=1,\ldots,P$. This collection of processes
defines a multivariate spatial process for composition. The $\alpha_{p}(s)$
processes are defined on the 8 km grid, $\alpha_{p}=\{\alpha_{p}(s_{1}),\ldots,\alpha_{p}(s_{m})\}$
for the $m$ grid cells. In \emph{Latent variable model}
we introduce a multinomial probit model that relates the $\alpha_{p}(s)$
processes to the proportion processes, $\theta_{p}(s)$, via the introduction
of latent variables, with an implicit sum-to-one constraint, $\sum_{p=1}^{P}\theta_{p}(s)=1$.

@@

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \cite{Bane:etal:2004} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with Matern
covariance \cite{Lind:etal:2011}.

\paragraph*{Spatial process models}

MRF models work directly with the precision matrix of the values of
the spatial process, so calculation of the prior density of $\alpha_{p}$
is computationally simple \cite{Rue:Held:2005}, but in situations
where the likelihood is not normal, it can be difficult to set up
effective MCMC algorithms that are able to move in the high-dimensional
space of $\alpha_{p}$. The latent variable representation helps to alleviate this problem.
Next we describe two alternative spatial models that we considered;
in \emph{Model comparison}, we use cross-validation to
choose between these two representations.

@@

Gaussian processes (GP) are also standard models for spatial processes \cite{Bane:etal:2004}.
GP models are computationally challenging for large datasets because
of manipulations involving large covariance matrices. Given this,
\cite{Lind:etal:2011} proposed a new framework for using Gaussian
MRFs (GMRFs) as approximations to GPs, based on the use of stochastic
partial differential equations (SPDEs).
